Structuring College Access: College Board Geo-markets and the Market Segment Model

### 2024-07-08 text/old_spring2024/background_em_long_theory_5_30_2024.qmd
initial commit, copying files from messed up repo

---
title: "Structuring College Access: The Market Segment Model and College Board Geo-markets"
author: ["Ozan Jaquette (UCLA)", "Karina Salazar (University of Arizona)"]
bibliography: ./../assets/bib/cb_geomarket_bib.bib
csl: ./../assets/bib/apa.csl
format:
  pdf:
    documentclass: article
    #papersize: letter
    number-sections: true
    number-depth: 2
    #toc: FALSE
    #pandoc_args: !expr rmdfiltr::add_wordcount_filter(rmdfiltr::add_citeproc_filter(args = NULL))
    #keep_tex: true
    extensions: [andrewheiss/wordcount]
editor: visual
geometry:
  margin=1in
fontsize: 11pt
header-includes:
  - \usepackage{setspace}
  - \doublespacing
---

@@

# https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_pdf.pdf
library(kableExtra)

#hey ho This many words: {{< words-total >}}
```

```{=tex}
\pagenumbering{roman}
\newpage
\pagenumbering{gobble}
\pagenumbering{arabic}
```

@@

In 1983, Zemsky and Oedel authored *The Structure of College Choice*, published by *The College Board.* This book develops the Market Segment Model and introduces local "community-based enrollment markets" (p.14) -- later called "geomarkets" -- which are the basis for the College Board market Enrollment Planning Service (EPS), a market research product. @RN4982[pp., 9-10] write that the Market Segment Model sought to quantify the knowledge that college admissions officers have about student demand:

> Admissions officers invariably are tellers of stories -- about the colleges they represent, about the colleges they attended, about each other, and about the often vagabond life of college recruiting\ldots We believe that the institutions of admissions officers actually comprise a remarkably systematic body of knowledge about the college selection process\ldots Our research, thus, is based on listening carefully to what admissions officers have to say."

The Market Segment Model was created by analyzing College Board data about which colleges SAT test-takers from 1980 sent scores to. High school students who sent scores mostly to colleges in their local geomarket were defined as "local" students. "In-state" students were those who sent most scores to colleges outside their local market but in their home state. "Regional" students were those who sent most scores to colleges outside their home state but in their region (e.g., New England, Midwest). "National" students were those who sent most scores to colleges outside their region. Having categorized students, each local geomarket could be evaluated by the number of students from each of the four market segments, yielding the Market Segment Profile (see @tbl-market-segment-characteristics). Next, for each local geomarket, the Institutional Profile (see @tbl-sample-institutional-profile) the number of students from each market segment that sent scores to a particular college, which could be your college or a competitor.

Collectively, the quantitative outputs of the Market Segment Model provided intelligence about where are the desirable students and which students might consider your college or a competitor. The Market Segment model and geomarkets became the basis or the College Board Enrollment Planning Service (EPS), founded in 1984 and still active today. @RN5019_2010 says "Enrollment Planning Service (EPS\textsuperscript{\textregistered}) the analysis tool that pinpoints the schools and geomarkets where your best prospects are most likely to be found." @RN5021 reports that in 1995, 37% of 4-year publics and 49% of 4-year privates used EPS, while 41% of 4-year publics and 16% of 4-year privates used ACT's market analysis service product, which was similar to EPS.

In 19XX, geomarkets were incorporated into the College Board Student Search Service student list product, which College Board created in 1972 [@belkin2019-studata]. Student lists contain the contact information of prospective students, the basic input for recruiting campaigns that target individual prospects using mail, email, and targeted social media. RN4895 reported that 87% of private and 86% of public four-year institutions purchase student lists. The Student Search Service database consists of College Board test-takers. Colleges pay a fee for each prospect (e.g., \$0.50 in 20XX) and control which prospect profiles through purchase by selecting search filters, such as high school graduation year, SAT score, AP test score, state, etc. Geomarket filters enable colleges to include/exclude prospects from particular geomarkets.

Scholarship in the sociology of education describes schools and colleges as "sorting machines" that reproduce inequality by incorporating, creating, and allocating students to different categories [@RN4873]. Given the consequences subsequent outcomes, the sociology of education is particularly concerned with the transition from high school to college. Over the last decade or so, scholarship on college access has begun to take more seriously the idea that colleges are not passive recipients of applications; rather they expend substantial resources soliciting demand from desirable applicants [@RN3519; @RN4774; @RN4324]. Most scholarship on recruiting analyzes the behavior of individual colleges, thereby assuming that recruiting behavior is a function of individual colleges [e.g., @RN4758; @RN4759]. @RN5012 analyze the College Board *Student Search Service* product, highlighting the role that third-party products play in structuring the recruiting behavior of colleges. The authors show that the utilization of particular search filters -- particularly geographic filters, such as zip code -- is associated with systematic racial exclusion that does not depend on which college purchased the list.

However, prior research has not examined the role of third-party market research in reproducing socioeconomic and racial inequality. This manuscript analyzes the Market Segment Model and College Board Geomarkets as a case study of quantification. We develop arguments from the broader literature on quantification [e.g.,] [@RN5003; @RN5005; @RN5007], particularly the discussions of correlation and homophily from @RN4975. Predictive analytics are based on correlation. When we use data on past correlations to make recommendations about the future, we amplify the effects of historic structural inequality. The Market Segment Model analyzes SAT score-sending data from 1980 and concludes that student demand for higher education is primarily a function of social class= [@RN4982]. Homophily is the idea that actors that share common characteristics are likely to form connections with one another. The Market Segment Model argues that homophily is the organizing principle of student demand and competition between colleges, that "the hierarchical structure of collegiate competition largely reflects the stratified social and economic dimensions of the communities from which colleges draw their students" [@RN4982, p. 72]. Subsequently, the assumption of homophily is programmed into the EPS and Student Service Service products that colleges utilize to decide where to focus recruiting energy.

Empirically, our analyses address the following research questions, which speak to how geomarkets are utilized within the Enrollment Planning Service and within the Student Search Service student list product. First, what is the socioeconomic and racial variation between geomarkets in metropolitan areas and how does this variation change over time? We address this question by spatially joining the geomarket shapefile to Census data about socioeconomic and racial characteristics. Second, how does the socioeconomic and racial composition of included versus included prospects vary when student list purchases filter on particular geomarkets, in concert with other commonly used search filters? We address this question by analyzing student list purchases that do not filter geomarkets and simulating which prospects would have been included had they filtered on particular geomarkets. We obtained these data by issuing public records requests to public universities in four states. Third, in student list purchases that filtered on geomarkets, how does the socioeconomic and racial composition of included students compare the that of the geomarket and how does it compare to that of the broader metropolitan area?

Scholarship on quantification in education has focused on consumer-facing measures such as U.S. News Rankings [@RN5003] and UK "league tables" about school quality [@RN5005]. To our knowledge, prior knowledge has not examined business-facing measures produced by third-party market research. Here, the desirable attributes of students and places they live are quantified with the purpose of informing colleges about where they should allocate recruiting resources to efficiently enroll the most desirable students. We reason that market research reproduces inequality by encouraging colleges to plan their recruiting efforts based on a snapshot of social stratification in student demand that is substantially a function of historic, structural inequality in educational opportunity. In contrast to @RN5005, we cannot demonstrate the effect of geomarkets on inequality. However, we show the mechanisms of how geomarkets reproduce inequality by simulating who is included/excluded when geomarkets are utilized within student list purchases as recommended by @RN4982.

In the following section we provide background information about enrollment management and scholarship on recruiting. Second, we introduce core ideas and concepts from scholarship on quantification. Third, we apply these ideas to a close read of @RN4982 and motivate hypotheses. Four, we describe data and methods. Fifth, we present results. Finally, we discuss implications for scholarship, policy, and practice.

# Background: Enrollment Management

@@

@@

Scholarship at the nexus of enrollment management and college access can be categorized by which part(s) of the enrollment funnel they focus on. The majority of scholarship focuses on the admissions stage, analyzing which admissions criteria are utilized and/or which applicants are admitted [e.g., @RN4874; @RN4993; @RN4131; @RN3522; @poon2022rethinking]. A growing literature analyzes the earlier "recruiting" stages of identifying leads, soliciting inquiries, and soliciting applications. Scholarship on recruiting has conceptualized recruiting behavior as an indicator of college enrollment priorities [@RN4758] and identifies recruiting practices as a mechanism for social reproduction in college access [@RN4324; @RN3519].

@@

Other studies investigate recruiting behavior by non-selective institutions [@RN4520; @cottom2017lower; @dache2018dangling]. Analyzing recruiting from the perspective of high school students, @RN4324 finds that underrepresented students were drawn to colleges that made them feel wanted, often attending institutions with lower graduation rates and requiring larger loans than other college options. @cottom2017lower shows that for-profit colleges found a niche in Black and Latinx communities because traditional colleges ignored these communities. <!--  WHAT YOU SAID ABOUT POSECZNICK Posecznick (2017) studied the “Ravenwood College,” a private nonprofit college founded to serve working Black women. Consistent with analyses of tuition discounting by Lapovsky (2001), the survival of Ravenwood depends on the success of admissions officers in identifying prospects (e.g., through on- and off-campus recruiting events) and “converting” these prospects to enrolled students.  -->

We make two observations about the recruiting literature. First, consistent with scholarship on political economy [@RN4517; @cottom2017lower; @RN5015] and industrial organization [@RN1549; @RN2133], the industry is characterized by a vertical hierarchy of postsecondary institutions matching to a vertical hierarchy of customer niches, a process that exemplifies the idea that higher education is mechanism of social reproduction [@RN2377; @RN4873].<!--  For example, for-profit institutions target adults in communities not being targeted by traditional colleges [@cottom2017lower]. Non-selective private non-profits target prospects who might be considering community college or a local state colleges [@RN4324]. Selective private colleges target private high schools, with a particular link between sectarian high schools and colleges of the same faith [@RN5014], but these schools are increasingly targeted by out-of-state public research universities, suggesting growing competition for a coveted market niche. --> Second, by focusing on the behaviors of individual colleges, the literature implicitly states that recruiting is something done by individual colleges. 
<!--  
Why then, does recruiting behavior look so similar across colleges of a particular "type"?[^1]
[^1]: New institutional theory offers one explanation [@RN527]. In highly developed organizational fields like higher education, forces of competitive and institutional isomorphism -- including institutional locics -- compel organizations adopt similar practices as their peers. Practices deemed advantageous spread quickly, for example, public research universities targeting nonresident prospects to compensate for declines in state appropriations. These behaviors are conceived as autonomous decisions by individual colleges, albeit structured by institutional logics.
 -->

<!-- make or buy  
The process of recruiting students involves many "make or buy" decisions about whether to perform a given task in-house or outsource it to a third-party vendor. EM consulting firms provide advice and implementation in the areas of marketing, recruiting, pricing and financial aid, and student success. As "creating a class" has becomes complicated and high-stakes [@RN3519], many colleges hire EM consulting firms to develop and/or implement recruiting campaigns. The two largest enrollment consulting firms -- Ruffalo Noel Levitz and EAB -- claim to serve more than 3,000 colleges and universities collectively [@rnlabout; @eabcomm]. In our data collection, attempting to obtain data about student list purchases from all public universities in four states, at least half of these universities outsourced student list purchases to an EM consultancy [@list_biz; @list_empirics].
-->

However, the recruiting behavior of colleges is structured by third-party actors. Drawing from scholarship on organizational theory [@RN4738], enrollment management processes involve many "make or buy" decisions about whether to perform a given task in-house or outsource it to a third-party vendor [@list_biz]. EM consulting firms provide advice and implementation in the areas of marketing, recruiting, pricing and financial aid, and student success. <!-- As "creating a class" has becomes complicated and high-stakes [@RN3519], many colleges hire EM consulting firms to develop and/or implement recruiting campaigns. --> The two largest enrollment consulting firms -- Ruffalo Noel Levitz and EAB -- claim to serve more than 3,000 colleges and universities collectively [@rnlabout; @eabcomm].

Moreover, process of college access is structured by third-party recruiting products utilized by colleges. @RN5012 deconstruct the College Board *Student Search Service* product. Selection devices are procedures or routines for making categorical decisions based on input values [@RN4778]. Student list products are selection devices that enable colleges to choose which prospective students they target or ignore based on search filters that incorporate search filters (e.g., high school graduating class, state) which enable colleges to select prospective students. "Racialized inputs" are ostensibly race-neutral inputs that are systematically correlated with race because marginalized racial/ethnic groups have historically been excluded from the input [@RN4786]. @RN5012 argue that several frequently utilized student list filters (zip code, AP test score, SAT score) are racialized inputs. Using a national sample of high school students, and also using data from actual student lists purchased by public universities, they show that racialized search filters have a strong negative relationship with the selection of Black and Latinx prospects. 
 <!-- Using data from actual student lists purchased by public universities, they show that raciaized search filters -- used in conjunction with other search filters -- yield dramatic racial inequality in which prospective students are targeted. -->

Student list products offer many geographic filters. Geographic borders are perhaps the most commonly studied racialized input in scholarship about algorithmic bias [@RN4775; @RN4849; @RN4887], owing to extreme racial segregation in the U.S. In addition to offering geographic filters based on known geographic borders (e.g., zip code, county, CBSA, state), the College Board created new geographic borders that subsequently became filters in the *Student Search Service*. The geodemographic segment filters utilize cluster analysis to allocate each census tract and each high school into different categories based past college enrollment and other factors [@RN4565]. Geodemographic search filters can be read as the mimicry of corporate market logic [@RN4004] in that they apply the logic and methods of commercial geodemographic market segmentation products, such as PRIZM by the Claritas Corporation [@RN4957], to the case of college recruiting.

Geomarket search filters, the focus of this manuscript, slice states and metropolitan areas into smaller, local recruiting markets. For example, geomarket "CA08" refers to "Alameda Country excluding Oakland." A university might purchase prospect contact information by filtering on this geographic region in combination with additional filters, such as year of high school graduation and SAT test score. When we initiated this manuscript, we asked ourselves, "where did geomarket filters come from?" The answer surprised us. We had imagined ourselves studying the application of the "data science revolution" to college recruiting, but it turned out we were studying the 1970s and 1980s! The Market Segment Model and geomarkets can be read as the application of social science research methods to "capture and quantify the phenomenon behind the folklore" [@RN4982, p. 11] of college admissions officers. This work, comissioned by College Board, was subsequently turned into a market research tool -- The College Board Enrollment Planning Service -- that helps colleges decide where to look for prospective students.

```{=html}
<!--The enrollment management industry emphasizes market research.
Market research is the process of producing information that informs decisions about product design or how to find customers.   -->
```

The enrollment management industry puts great stock in market research. @RN4986[p. 17] define market research as all "activities intended to increase understanding of a market and consumer behavior in that market." At the most basic level, in-house market research is a function of being intentional about record-keeping and using those records [@RN3519].^[@RN3519 describes the archives of annual reports of admissions activities from a selective liberal arts college. In the 1942 report, "Section 2.5 is titled 'Secondary School Quartile Standings," a measure of where matriculants fell in the rankings of their high school classes" [@RN3519] while Section 3.11 describes the high schools visited. In the chapter entitled "Travel," @RN3519 [p. 69] describes how The College used record-keeping to inform decisions about visits to local high schools in the 2000s: "The College keeps files on hundreds of schools across the country, and around the globe, with which it had contact \ldots The files enable officers to quickly discern"how we've done here lately," \ldots with the number, relative quality, and yield rate of applicants from a particular school."] However, most market research in the contemporary enrollment management industry is produced by third-party organizations. For example, Ruffalo-Noel Levitz and EAB both produce prodigious amounts of market research, some publicly available as a means of business development [e.g., @RN4741; @RN4728, but most only available to clients. EPS remains the market research arm of College Board, while Encoura Eduventures is the market research arm of ACT.

Despite being central to enrollment management for at least 40 years [@RN4986], the role of third-party market research has been ignored by scholarship on college access. Scholarship on college access often takes a demand-side perspective, whereby household differences in economic and cultural capital yield stratification in student preferences and the ability to gain admission [e.g., @RN4935;@RN1814]. However, market research informs where colleges look for students and which prospective students to target. These recruiting interventions intervene on student demand [@RN4324; @cottom2017lower], thereby structuring college access pathways in ways that could amplify or diminish historic inequality in college access. The College Board played a pivotal role in transforming market research about enrollment from something done by individual universities to something done by third-party professionals [@RN4986]. This article offers one story about the quiet revolution that transformed college access into a sophisticated, supply-led enterprise. We argue that market research structures who colleges recruit, in ways that reinforce existing social stratification. The case of Market Segment Model and College Board geomarkets shows how market research -- and its commodification into recruiting products -- is a mechanism for social reproduction. 
[LAST TWO SENTENCES ARE REDUNDANT BUT BUT ARE TRYING TO DO SLIGHTLY DIFFERENT THINGS AND I LIKE THEM BOTH]

@@

Scholarship about U.S. News & World Report (USNWR) Law School rankings demonstrate the reactivity, discipline, and authority effects of quantification [@RN5003; @RN2356; @RN2376; @RN2375]. Developed as a means of informing prospective students about quality, rankings had the intended effect of affecting. Rankings also affected hiring decisions of firms because clients evaluated firms based on the prestige of law schools attended by firm lawyers. Once law schools realized that "important groups of constituents --- students, faculty, trustees, employers, other media --- were using rankings to make decisions that had large consequences for schools \ldots \[then\] schools felt pressured to take them seriously" [@RN5007, p. 415]. Rankings *disciplined* the behavior of law schools. For example, applicants with characteristics valued by the rankings system (e.g., LSAT scores) became more important for decisions about admissions and merit aid [@RN5003]. Rankings also weakened the *authority* of admissions directors and admissions officers to make independent decisions about which applicants to admit, ideal class size, and curricular offerings [@RN5003]. <!--
In a process akin to sports analytics in Moneyball \[CITE MICHAEL LEWIS\], we suggest that market research replaces the wisdom and authority of local admissions counselors.    -->

The analysis of UK League Tables by @RN5005 show how quantification can be a mechanism of social reproduction. As part of a broader *New Public Management* effort to infuse market competition within public providers [@RN161], the UK government began publishing school "league tables" in 1992. League tables ranked schools based on student performance on high-stakes national exams taken at age 16. Using Census data on occupational class measured at the local authority level (similar to a U.S. county), @RN5005 find that localities with higher performing schools experienced growth in the share of managerial/professional households following the adoption of league tables and a decline in the share of working-class households. Additionally, using longitudinal household survey data, managerial/professional households were more likely to move to localities with higher ranked schools following the introduction of school league tables. With respect to mechanisms, league tables facilitate making evaluative comparisons between schools regardless of geographic proximity. They also discipline households and schools to conceive of school quality in terms of test scores, which are substantially a function of the class composition of schools. Compared to working-class parents, parents with professional/managerial occupations are more aware of school league tables and have financial resources necessary to move to neighborhoods near a "good" school. Therefore, league tables contributed to social reproduction because managerial/professional households respond to consumer-facing metrics by moving to more expensive neighborhoods, near higher performing schools. <!-- THIS IS RATCHET EFFECT STUFF. I CUT IT BECAUSE YOU HAVEN'T INTRODUCED THE IDEA OF RATCHET EFFECT
Schools that performed well on initial league tables tended to be schools with a high share of professional/managerial families. Therefore, quantification takes a snapshot of existing social stratification and that snapshot becomes a power unto itself, amplifying the effect of previous strctural inequality on subsequent structural inequality.
-->

```{=html}
<!-- THIS IS THE CUT TRANSITION PARAGRAPH
In the broader quantification literature, critical data studies [e.g., @RN4975; @RN4849; @RN4772] and sociological analyses of technology and inequality [e.g., @RN4778; @RN4786; @RN4877; @RN4810] introduce concepts salient to the analysis of market research. Our focus is on "selection devices," which utilize some combination of inputs and discretion to categorize places, people, and organizations for the purpose of allocating opportunities or interventions [@RN4778; @RN5012].
-->
```
In the interdisciplinary literature on quantification [@RN5008], the discussions of correlation and homophily by @RN4975 introduce concepts that are salient to the analysis of market research. Chapter 1 of @RN4975 is *Correlating Eugenics*. Correlation measures the extent to which two or more variables move together. The Pearson Correlation Coefficient was developed by the 19th Century statisticians and "biometric eugenicists" Francis Galton and his student Karl Pearson [@RN4975]. @RN4975 [p. 50] writes that "correlation grounds big data's so-called revolutionary potential. As Wired editor Chris Anderson infamously declared in his 2008 editorial 'The End of Theory,' big data proved that 'correlation supersedes causation, and science can advance even without coherent models, unified theories, or really any mechanistic explanation at all.'" Predictive analytics are developed in two steps. First, apply statistical techniques to previous cases (training data) in order to identify factors positively and negatively associated with an outcome of interest. Second, apply these results (e.g., regression coefficients) to future cases in order to make predictions and to assign levels of risk to each case. These models predict outcomes based on correlations without requiring knowledge about underlying causal relationships. @RN4975 provides the example of @RN4989, who develop a method to predict sensitive personal attributes (e.g., gender, political party) based on Facebook Likes.

>>>>

---
title: "Structuring College Access: College Board Geo-markets and the Market Segment Model"
author: ["Ozan Jaquette (UCLA)", "Karina Salazar (University of Arizona)"]
bibliography: ./../assets/bib/cb_geomarket_bib.bib
csl: ./../assets/bib/apa.csl
format:
  pdf:
    documentclass: article
    #papersize: letter
    number-sections: true
    number-depth: 2
    #toc: FALSE
    #pandoc_args: !expr rmdfiltr::add_wordcount_filter(rmdfiltr::add_citeproc_filter(args = NULL))
    #keep_tex: true
editor: visual
geometry:
  margin=1in
fontsize: 11pt
header-includes:
  - \usepackage{setspace}
  - \doublespacing
---

@@

# https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_pdf.pdf
library(kableExtra)
```

@@

hey ho This many words: {{< words-total >}}

COULD START WITH HISTORICAL VIGNETTE OF ZEMSKY PARTNERING WITH COLLEGE BOARD TO CREATE THE MARKET SEGMENT MODEL.

-   quotes on "admissions officers are invariably tellers of stories" and the goal of the market segment model is to quantify these stories.

TEXT TEXT

Thinking about a revised outline, 4/19/2024

-   Intro/lit review
-   theory \[e.g., quantification, whatever\]
-   then do a close read of the zemsky book, which ends in creation of EPS, incorporating geomarkets in student list products
    -   hypotheses would be motivated here
-   methods/data

the alternative (typical education journal) structure would be this \[which has one more level 1 heading and feels to inefficient\]:

-   introduction \[topic importance\]
-   literature review
-   background
-   conceptual framework
-   methods data

IMPORTANT (defense/offense): One important limitation of this manuscript is that we do not study the effect of geomarkets on college access. However, we do show empirically how geomarkets work as a mechanism of exclusion and social reproduction.

Prior research on quantification in education focuses on consumer-facing products. For example, Espeland and Sauder study US News Rankings. McArthur, D., & Reeves, A. (2022) study School League Tables. Prior research has not studied quantification within the context of market research. Here, the desirable attributes of students and places they live are quantified with the purpose of informing colleges about where they should allocate recruiting resources to efficiently enroll the most desirable students.

```{=tex}
\pagenumbering{roman}
\newpage
\pagenumbering{gobble}
\pagenumbering{arabic}
```
# Background: Enrollment Management

@@

The process of recruiting students involves many "make or buy" decisions about whether to perform a given task in-house or outsource it to a third-party vendor. EM consulting firms provide advice and implementation in the areas of marketing, recruiting, pricing and financial aid, and student success. As "creating a class" has becomes complicated and high-stakes [@RN3519], many colleges hire EM consulting firms to develop and/or implement recruiting campaigns. The two largest enrollment consulting firms -- Ruffalo Noel Levitz and EAB -- claim to serve more than 3,000 colleges and universities collectively [@rnlabout; @eabcomm]. In our data collection, attempting to obtain data about student list purchases from all public universities in four states, at least half of these universities outsourced student list purchases to an EM consultancy [@list_biz; @list_empirics].

@@

Scholarship at the nexus of enrollment management and college access can be categorized by which part(s) of the enrollment funnel they focus on. The majority of scholarship focuses on the admissions stage, for example which admissions criteria are used [@RN4874; @RN4993] and which applicants are admitted [@RN4131; @RN3522] \[CITE SOMETHING BY BASTEDO\].

A growing literature analyzes the earlier "recruiting" stages of identifying leads, soliciting inquiries, and soliciting applications. Scholarship on recruiting has conceptualized recruiting behavior as an indicator of college enrollment priorities [@RN4758] and identifies recruiting practices as a mechanism for social reproduction in college access [@RN4324; @RN3519].

@@

Other studies have investigated recruiting behavior by non-selective institutions [@RN4520; @cottom2017lower]. Analyzing recruiting from the perspective of high school students, @RN4324 finds that underrepresented students were drawn to colleges that made them feel wanted, often attending institutions with lower graduation rates and requiring larger loans than other college options. @cottom2017lower and @dache2018dangling show that for-profit colleges found a niche in Black and Hispanic communities because traditional colleges ignored these communities. <!--  WHAT YOU SAID ABOUT POSECZNICK Posecznick (2017) studied the “Ravenwood College,” a private nonprofit college founded to serve working Black women. Consistent with analyses of tuition discounting by Lapovsky (2001), the survival of Ravenwood depends on the success of admissions officers in identifying prospects (e.g., through on- and off-campus recruiting events) and “converting” these prospects to enrolled students.  -->

We make two observations about the recruiting literature. First, consistent with scholarship on political economy [@RN4517; @cottom2017lower] \[cite Stevens; cottom\] and industrial organization \[cite Winston\], the recruiting literature suggests a hierarchical competition in which different "types" of postsecondary institutions target hierarchically arranged customer niches. For example, for-profit institutions target adults in communities not being targeted by traditional colleges [@cottom2017lower]. Non-selective private non-profits target prospects who might be considering community college or a local state colleges [@RN4324]. Selective private colleges target private high schools, with a particular link between sectarian high schools and colleges of the same faith [@RN5014], but these schools are increasingly targeted by out-of-state public research universities, suggesting growing competition for a coveted market niche. This process of hierarchically organized colleges matching to hierarchical student segments exemplifies the idea that higher education is mechanism of social reproduction [@RN2377; @RN4873]

Second, by focusing on the behaviors of individual colleges, the literature makes the implicit assumption that recruiting is something done by individual colleges. Why then, does recruiting behavior look so similar across colleges of a particular "type"?[^1]

[^1]: New institutional theory offers one explanation [@RN527]. In highly developed organizational fields like higher education, forces of competitive and institutional isomorphism -- including institutional locics -- compel organizations adopt similar practices as their peers. Practices deemed advantageous spread quickly, for example, public research universities targeting nonresident prospects to compensate for declines in state appropriations. These behaviors are conceived as autonomous decisions by individual colleges, albeit structured by institutional logics.<!--WHAT DOMINA (2017) SAID ABOUT LOGICS, P. 322: Like other distinctively modern institutions, contemporary schools are governed by standardized and rationalized institutional logics (Baker 2014, Bowker & Star 2000, Timmermans & Epstein 2010, Meyer 1977). These logics have profound implications for educational category assignment processes, leading educators, policy makers, scholars, and others to debate the universality of selection-->

@RN5012 argues that the process of college access is structured by third-party recruiting products utilized by colleges. @RN5012 deconstructs the College Board *Student Search Service* product. Student list products are "selection devices" [@RN4778] that incorporate search filters (e.g., high school graduating class, state) which enable colleges to select prospective students. Drawing from @RN4786, @RN5012 argue that several search filters (e.g., AP test score, zip code, geodemographic segment) are "racialized inputs," defined as ostensibly race-neutral inputs that are correlated with race because disadvantaged racial groups have historically been excluded from the input. Using a national sample of high school students from the High School Longitudinal Study of 2009 (HSLS), they show that racialized search filters have a strong negative relationship with the selection of Black and Latinx prospects. Using data from actual student lists purchased by public universities, they show that raciaized search filters -- used in conjunction with other search filters -- yield dramatic racial inequality in which prospective students are targeted.

[NEED CLEANER TRANSITION]Two geographic search filters -- geo-market and geodemographic segment -- that @RN5012 observe in lists purchased by public could not recreate using HSLS. Using geographic borders to determine selection for an intervention has been widely studied as a mechanism of structural inequality because of historic and contemporary racial segregation in American communities and schools [e.g., @RN4775; @RN4849; @RN4779]. The geo-market and geodemographic segment search filters of the College Board *Student Search Service* are examples of creating new geographic borders for the purpose of targeting students for educational opportunities. The geo-market filter slices states and metropolitan areas into smaller, "local" markets. For example, CA08 refers to "Alameda Country excluding Oakdland" and a university might purchase prospect contact information by filtering on this geographic region, alongside additional filters. The geodemographic segment filters utilize cluster analysis to allocate each census tract and each high school into different categories based past college enrollment and other factors. This manuscript focuses on geo-markets. The first question we asked ourselves was, "where did these geo-markets come from?" The answer is, market research.

```{=html}
<!--The enrollment management industry emphasizes market research.
Market research is the process of producing information that informs decisions about product design or how to find customers.   -->
```
The enrollment management industry puts great stock in market research. @RN4986 define market research as all "activities intended to increase understanding of a market and consumer behavior in that market" (p. 17). At the most basic level, (in-house) market research is a function of intentional record-keeping. @RN3519 describes the archives of annual reports of admissions activities from a selective liberal arts college. In the 1942 report, "Section 2.5 is titled 'Secondary School Quartile Standings," a measure of where matriculants fell in the rankings of their high school classes" [@RN3519] while Section 3.11 describes the high schools visited. In the chapter entitled "Travel," @RN3519 [p. 69] describes how The College used record-keeping to inform decisions about visits to local high schools in the 2000s: "The College keeps files on hundreds of schools across the country, and around the globe, with which it had contact \ldots The files enable officers to quickly discern"how we've done here lately," \ldots with the number, relative quality, and yield rate of applicants from a particular school."

The College Board played a pivotal role in transforming market research about enrollment from something done by individual universities to something done by third-party professionals. In the late 1970s, College Board provided Carleton College with funding and data for a six-year project that analyzed college choice decisions by households in six metropolitan areas.[^2] This project was the basis for the landmark book *Applying Market Research in College Admissions* [@RN4986], published by College Board, and was the basis for the College Board Admitted Student Questionnaire (ASQ), a revenue-generating product that provides colleges with information about admitted students -- those who enrolled and who did not enroll. <!--Historically, College Board partnered with Members to develop projects that later became revenue-generating products for College Board.    -->

[^2]: In the 1970s, Carleton College became a leader in applying market research to student recruiting, under the leadership of Larry Litten, the then director of institutional research. Litten would later become Associate Director of the Consortium of Financing Higher Education (COFHE).

Similarly, in 1979 The College Board provided Robert Zemsky, a professor of education at UPenn, with data and funding to develop the *Comprehensive Undergraduate Enrollment Planning Project* [@RN4982]. In 1983, The College Board published *The Structure of College Choice* by Robert Zemsky and Penney Oedel, developed the "market segment model" and defined local "geo-markets" based on a comprehensive analysis of SAT score-sending by prospective students to coleges and universities. In 1984/5, the market segment model and geo-markets became the basis for the College Board Enrollment Planning Service (EPS), the market research division of College Board.

This manuscript analyzes the market segment model and geo-markets as a case study in order to investigate market research -- and its commodification into recruiting products -- as a mechanism for social reproduction. At root, third-party market research helps universities identify who are the desirable students, where are they located, which institutions are they considering, and how best to reach them. Today, Ruffalo-Noel Levitz and EAB both produce prodigious amounts of market research, some publicly available as a means of business development \[e.g., CITE RNL\], but most only available to clients. EPS remains the market research arm of College Board, while Encoura Eduventures is the market research arm of ACT. Despite being central to enrollment management for at least 40 years [@RN4986], the role of third-party market research has been ignored by scholarship on college access, which views enrollment management behavior as the autonomous decisions of individual colleges, albeit structured by institutional logics [@RN4758; @RN4873; @RN3519]. We argue that market research structures who colleges recruit, in ways that reinforce existing social stratification.

@@

Scholarship about U.S. News & World Report (USNWR) Law School rankings demonstrate the reactivity, discipline, and authority effects of quantification [@RN5003; @RN2356; @RN2376; @RN2375]. Developed as a means of informing prospective students about quality, rankings affected which law schools applicants applied to, the intended effect. However, rankings also affected hiring decisions of firms because clients evaluated firms based on the prestige of law schools attended by firm lawyers. (clients evaluated firms based on the prestige of the law schools that firm lawyers attended) and

USNWR developed the rankings as a means to inform prospective students about law school quality. However, rankings affected

US News developed law schools rankings as a means of informing prospective students about law school quality. In turn, rankings affected how prospective students conceive law school quality and which schools they applied to. Rankings also affected the cognition and behavior of other actors. Looking downstream, entities that received law school students as an input -- for example, law firms -- used rankings as a hiring criterion because their clients evaluated the organization based on the the prestige of the schools their firm's lawyers attended.

Drawing from Foucault \[CITE DISCIPLINE AND PUNISH\], discipline is the idea that surveilance -- attributes of a law school being quantified and aggregated into a public-facing metric -- compels people and organizations to react in a particular way. For example, the reputational assessment of peer leaders as an important component of most US News rankings. Bastedo and Bowman showed that once (undergraduate) USNWR rankings became popular, the reputational assessment of peers primarily became a function of previous USNWR rankings. Once law schools realized that "important groups of constituents --- students, faculty, trustees, employers, other media --- were using rankings to make decisions that had large consequences for schools \ldots schools felt pressured to take them seriously" \[QUOTE FROM ESPELAND AND STEVENS\]. For example, applicants with characteristics valued by the rankings system (e.g., LSAT scores) became more important for decisions about admissions and merit aid \[CITE\].

The theme of authority is that quantification destabilizes and weakens the authority of local decision-makers. For example, law school rankings weakened the authority of law schools to make independent decisions about which applicants to admit, ideal class size, and curricular offerings \[CITE THE BOOK\]. In a process akin to sports analytics in Moneyball \[CITE MICHAEL LEWIS\], we suggest that market research replaces the wisdom and authority of local admissions counselors.

The analysis of UK League Tables by @RN5005 show how quantification can be a mechanism of social reproduction. In 1992, as part of a the broader *New Public Management* effort to instill market competition amongst public providers [@RN161], the UK government started publishing school "league tables" that ranked schools based on student performance on the high-stakes General Certificate of Secondary Education (GCSE) exams that students take at age 16. @RN5005 argue that quantifying the quality of schools entrenches class-based inequalities between schools particularly when the measure of quality -- standardized test scores -- is substantially a function of the class composition of schools. School league tables discipline households and schools to conceive of school quality in terms of test scores and they facilitate making evaluative comparisons between schools that are geographically close or distant. Compared to working-clas parents, parents with professional/managerial occupations are more aware of school league tables and have financial resources necessary to move to neighborhoods near a "good" school.

Using school-level data and data on census data on occupational class measured at the local authority level (similar to a U.S. county), @RN5005 find that, following the adoption of league tables, local areas with higher performing schools experience growth in the share of managerial/professional occupational classes and a decline in the share of working class families. Additionally, using longitudinal survey data on 1% of the population of England and Wales and a diference-in-difference estimator, the authors find that managerial/professional households are more likely to move to local areas with higher ranked schools following the introduction of school league tables. Overall, @RN5005 show that quantifying which schools are high-performing contributes to social reproduction because managerial/professional households respond to consumer-facing metrics by moving to more expensive neighborhoods, near higher performing schools. Schools that performed well on initial league tables tended to be schools with a high share of professional/managerial families. Therefore, quantification takes a snapshot of existing social stratification and that snapshot becomes a power unto itself, amplifying the effect of previous strctural inequality on subsequent structural inequality.

In the broader quantification literature, critical data studies [e.g., @RN4975; @RN4849; @RN4772] and sociological analyses of technology and inequality [e.g., @RN4778; @RN4786; @RN4877; @RN4810] introduce concepts salient to the analysis of market research. Our focus is on "selection devices," which utilize some combination of inputs and discretion to categorize places, people, and organizations for the purpose of allocating opportunities or interventions [@RN4778; @RN5012].
